\documentclass{article}


% if you need to pass options to natbib, use, e.g.:
\PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2022


% ready for submission
\usepackage{neurips_2022}
\include{packages}
\include{macros}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2022}


% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2022}


% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2022}




\title{Fast Variational Estimation of Mutual Information for Implicit and Explicit Likelihood Models}


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.


\author{%
  Caleb Dahlke \\
  Department of Applied Math\\
  University of Arizona\\
  Tucson, USA \\
  \texttt{calebdahlke@math.arizona.edu} \\
  % examples of more authors
  \And
  Sue Zheng\\
  Analog Devices \\
  Cambridge, MA \\
  \texttt{szheng@csail.mit.edu} \\
  \AND
  Jason Pacheco \\
  Department of Computer Science \\
  University of Arizona\\
  ATucson, USA \\
  \texttt{pachecoj@cs.arizona.edu} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}


\begin{document}


\maketitle


\begin{abstract}
  Computing the mutual information (MI) between sets of random
  variables lacks a closed-form solution in nontrivial
  models. Variational MI approximations are widely used as flexible
  estimators for this purpose, but computing these estimators
  typically requires solving a costly non-convex optimization.  In
  this paper we demonstrate a class of variational estimators whose
  solution is not only convex, but efficiently solved via moment
  matching conditions.  We show that the moment-matched variational
  estimator provides optimal upper and lower bounds on MI in models
  with explicit forward models.  Furthermore, we show that the same
  moment matching solution yields accurate MI approximations in
  so-called "implicit likelihood models", where the observation
  likelihood lacks an analytic representation.  In further analysis on
  implicit likelihood models we prove that our moment matching
  solution is equivalent to existing gradient based methods but at a
  significantly reduced computational cost. Our theoretical results
  are supported by numerical evaluation, showing that the proposed
  approach elegantly applies to fully parameterized (explicit
  likelihood) Gaussian mixture models, as well as implicit models
  arising from marginalization of nuisance variables.  Finally, using
  the SIR model in epidemiology, we show that our approach easily
  applies to implicit simulation-based likelihood models, while
  avoiding costly Likelihood-Free Inference Ratio Estimation (LFIRE)
  common to such models.
\end{abstract}

%% OLD ABSTRACT %%
%% \begin{abstract}
%% Computing the mutual information (MI) between sets of random variables
%% lacks a closed-form in nontrivial models.  Variational approximations
%% provide an efficient mechanism for estimating MI in complex models.
%% However, the presence of nuisance variables results in an implicit
%% relationship between the variables of interest, thus prohibiting the
%% use of most variational MI approximations.  This so-called "implicit
%% likelihood" model is supported by specialized variational
%% approximations, but they can be impractical in realistic settings
%% since they require expensive gradient-based updates.  In this paper we
%% show loose conditions whereby variational approximations over implicit
%% likelihood models can be efficiently solved by simple moment matching
%% operations, thus avoiding the need for costly gradient optimization.
%% Moreover, we prove that moment matching finds the identical solution
%% to existing optimization methods at a significantly reduced
%% computational cost.  We also provide conditions whereby our approach
%% achieves lower approximation error compared to widely used variational
%% MI approximations, even when the likelihood is not implicit.  Our
%% theoretical results are supported by numerical evaluation involving
%% A/B statistical hypothesis testing and nonlinear sensor selection
%% state-space models.
%% \end{abstract}

\section{Introduction}\label{Intro}
\input{introduction}

\section{Computing and Approximating MI}\label{background}
\input{background}

\section{Variational MI Estimation}\label{varapprox}
\input{varapprox}

\section{Moment Matching Variational MI Estimators}\label{varoptim}
\input{varoptimization}

\section{Previous Work}\label{previousWork}
\input{prev}

\section{Experiments}\label{experiments}
\input{experiments}

\section{Discussion}\label{discussion}
\input{discussion}


\bibliography{refs}
\bibliographystyle{abbrvnat}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Checklist}


%%% BEGIN INSTRUCTIONS %%%
The checklist follows the references.  Please
read the checklist guidelines carefully for information on how to answer these
questions.  For each question, change the default \answerTODO{} to \answerYes{},
\answerNo{}, or \answerNA{}.  You are strongly encouraged to include a {\bf
justification to your answer}, either by referencing the appropriate section of
your paper or providing a brief inline description.  For example:
\begin{itemize}
  \item Did you include the license to the code and datasets? \answerYes{See Section~\ref{gen_inst}.}
  \item Did you include the license to the code and datasets? \answerNo{The code and the data are proprietary.}
  \item Did you include the license to the code and datasets? \answerNA{}
\end{itemize}
Please do not modify the questions and only use the provided macros for your
answers.  Note that the Checklist section does not count towards the page
limit.  In your paper, please delete this instructions block and only keep the
Checklist section heading above along with the questions/answers below.
%%% END INSTRUCTIONS %%%


\begin{enumerate}


\item For all authors...
\begin{enumerate}
  \item Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
    \answerYes{}
  \item Did you describe the limitations of your work?
    \answerYes{}
  \item Did you discuss any potential negative societal impacts of your work?
    \answerNA{}
  \item Have you read the ethics review guidelines and ensured that your paper conforms to them?
    \answerNA{}
\end{enumerate}


\item If you are including theoretical results...
\begin{enumerate}
  \item Did you state the full set of assumptions of all theoretical results?
    \answerYes{}
        \item Did you include complete proofs of all theoretical results?
    \answerYes{Will be included in Apendix.}
\end{enumerate}


\item If you ran experiments...
\begin{enumerate}
  \item Did you include the code, data, and instructions needed to reproduce the main experimental results (either in the supplemental material or as a URL)?
    \answerYes{Will be included in Apendix}
  \item Did you specify all the training details (e.g., data splits, hyperparameters, how they were chosen)?
    \answerYes{}
        \item Did you report error bars (e.g., with respect to the random seed after running experiments multiple times)?
    \answerYes{}
        \item Did you include the total amount of compute and the type of resources used (e.g., type of GPUs, internal cluster, or cloud provider)?
    \answerYes{}
\end{enumerate}


\item If you are using existing assets (e.g., code, data, models) or curating/releasing new assets...
\begin{enumerate}
  \item If your work uses existing assets, did you cite the creators?
    \answerNA{}
  \item Did you mention the license of the assets?
    \answerNA{}
  \item Did you include any new assets either in the supplemental material or as a URL?
    \answerNo{}
  \item Did you discuss whether and how consent was obtained from people whose data you're using/curating?
    \answerNA{}
  \item Did you discuss whether the data you are using/curating contains personally identifiable information or offensive content?
    \answerNA{}
\end{enumerate}


\item If you used crowdsourcing or conducted research with human subjects...
\begin{enumerate}
  \item Did you include the full text of instructions given to participants and screenshots, if applicable?
    \answerNA{}
  \item Did you describe any potential participant risks, with links to Institutional Review Board (IRB) approvals, if applicable?
    \answerNA{}
  \item Did you include the estimated hourly wage paid to participants and the total amount spent on participant compensation?
    \answerNA{}
\end{enumerate}


\end{enumerate}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\appendix


\section{Appendix}
\input{appendix}


\end{document}
