To optimize the variational posterior lower bound, $\Ipost$, we must minimize the
bound gap.
\begin{equation}\label{eq:PostMin}
  I(X;Y) \geq \max_{q_p} \, H_p(X) - H_p(q_p(X \mid Y)) = \min_{q_p} \, H_p(q_p(X \mid Y)) + \text{const}
\end{equation}
The $H_p(X)$ term is constant in $q_p$ and can be ignored for
optimization. The maximization on the left is turned into a
minimization on the right by the negative leading the conditional
entropy.  \EQN\eqref{eq:PostMin} is optimized by satisfying the following
condition,
\begin{lemma}\label{lemma:post_opt}
  If $q_p(x\mid y)$ takes the form of \EQN\ref{eq:ExpFamCond}, then the minimization
  of \EQN\eqref{eq:PostMin} is when
  \begin{equation}\label{eq:OptCrit}
    \E_{p(y)}\left[\E_{q_p(x\mid y)}\left[T(X,Y)\right]\right]=\E_{p(x,y)}\left[T(X,Y)\right]
  \end{equation}
\end{lemma}\vspace*{-2mm}
This result was also shown previously by Pacheco and
Fisher~(\citeyear{pacheco2019variational}).  The condition in
Lemma~\ref{lemma:post_opt} is seemingly a moment matching condition.
However, the l.h.s.~of \EQN\ref{eq:OptCrit} is an expectation
w.r.t.~mixed distributions $p(y)q_p(x\mid y)$, and is difficult to satisfy
in general.  To simplify we consider the joint exponential family
distribution $q(x,y;\eta)$ with natural parameters $\eta$ and
conditional given by,
\begin{equation}\label{eq:ExpFamCond}
  q_p(x\mid y)=q(x\mid y;\eta)=\dfrac{q(x,y;\eta)}{q(y;\eta)}
\end{equation}\label{lemma:OptCrit}
Note that $q(y;\eta)=\int q(x,y;\eta) \, dx$ is not necessarily in the
exponential family.  We can now show that moment matching joint
statistics of $q(x,y;\eta)$ yields the optimal $\Ipost$ via the
following Lemma.
\begin{lemma}\label{lemma:MMOptCrit}
  Let $q_p(x\mid y)$ takes the form of \EQN\ref{eq:ExpFamCond}. Further,
  let the posterior expected statistics be a linear combination of
  marginal statistics as in,\vspace*{-3mm}
  \begin{equation}\label{eq:linearY}
    \E_{q_p(x\mid y)}\left[T(X,Y)\right]=\sum_i^k g_i(\eta)T_{i}(Y)
  \end{equation}
  where $T_i(y)$ is the $i^{th}$ component of the joint statistics
  only depending on  $y$ and $g_i(\eta)$ are arbitrary functions. 
  Then, the optimal $\Ipost$ is given by joint moment matching:
  $\E_{p(x,y)}[ T(X,Y) ] = \E_{q(x,y)}[ T(X,Y) ]$. 
\end{lemma}
Thus both of these conditions imply that moment matching the joint is
optimal for \EQN\eqref{eq:PostMin}. Each of these lemmas are written
in term of an exponential family distribution with some conditions,
since the multivariate Gaussian is the focus of this paper we need to
show that it satisfies these conditions
\begin{corollary}\label{cor:LinearGaussian}
  Let $q(x, y) = \Ncal(m, \Sigma)$ be a Gaussian. Then $q_p(x \mid y)$
  is also Gaussian and satisfies conditions of both
  lemma~\ref{lemma:OptCrit} and lemma~\ref{lemma:MMOptCrit}.
  Furthermore, the optimal $\Ipost$ is obtained by joint Gaussian
  moment matching conditions,\vspace*{-2mm}
  \[
    m^* = \E_{p(x,y)}\left[ (X,Y)^T \right], \qquad \Sigma^* = \text{Cov}_{p(x,y)}\left((X,Y)^T\right)
  \]
  \vspace*{-3mm}And moments of $q_p(x \mid y)$ are the corresponding Gaussian conditional moments
  of $m^*$ and $\Sigma^*$.
\end{corollary}
